---
title: "STAT 4310 Regression Project"
author: "Thomas Linden, Louis Mestas, Luis Garcia, Emanuel Padilla, Eugene Kazi, Edgar Pineda"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "project-presentation.css"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 220,
  fig.retina = 2,
  out.width = "100%"
)

library(AER)
library(MASS)
library(car)
library(lmtest)
library(ModelMetrics)
library(broom)
library(knitr)
library(quantmod)

set.seed(4310)

get_metrics <- function(actual, predicted) {
  data.frame(
    MSE = mse(predicted, actual),
    RMSE = rmse(predicted, actual),
    MAE = mae(predicted, actual),
    Test_R2 = 1 - sum((actual - predicted)^2) /
      sum((actual - mean(actual))^2)
  )
}

get_diag <- function(model) {
  data.frame(
    Test = c("Shapiro-Wilk", "Breusch-Pagan", "Durbin-Watson"),
    "P Value" = trimws(format.pval(c(
      shapiro.test(residuals(model))$p.value,
      bptest(model)$p.value,
      dwtest(model)$p.value
    ), digits = 3, eps = 1e-16))
  )
}

get_term_table <- function(model) {
  out <- tidy(model)
  out <- out[out$term != "(Intercept)", c("term", "estimate", "p.value")]
  out <- out[order(out$p.value), ]
  out$estimate <- sprintf("%.4f", out$estimate)
  out$p.value <- trimws(format.pval(out$p.value, digits = 3, eps = 1e-16))
  names(out) <- c("Term", "Estimate", "P Value")
  rownames(out) <- NULL
  out
}

format_test_rmse <- function(aic_step, cv_step, aic_rmse, cv_rmse) {
  if (identical(aic_step, cv_step)) {
    sprintf("%.2f (same for AIC and CV)", aic_rmse)
  } else {
    sprintf("AIC %.2f, CV %.2f", aic_rmse, cv_rmse)
  }
}

subgroup_r2 <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}

linear_combo_stats <- function(model, weights) {
  beta <- coef(model)
  v <- vcov(model)
  w <- rep(0, length(beta))
  names(w) <- names(beta)
  w[names(weights)] <- weights

  estimate <- sum(w * beta)
  se <- sqrt(drop(t(w) %*% v %*% w))
  df <- df.residual(model)
  crit <- qt(0.975, df)
  t.value <- estimate / se
  p.value <- 2 * pt(-abs(t.value), df)

  c(
    Estimate = estimate,
    Lower = estimate - crit * se,
    Upper = estimate + crit * se,
    P.Value = p.value
  )
}

build_backward_ladder <- function(model) {
  models <- list(model)
  steps <- data.frame(
    step_num = 0,
    step = "Step 0",
    predictors = length(attr(terms(model), "term.labels")),
    aic = AIC(model),
    removed = NA_character_,
    stringsAsFactors = FALSE
  )

  current_model <- model

  while (length(attr(terms(current_model), "term.labels")) > 1) {
    current_terms <- attr(terms(current_model), "term.labels")

    candidate_models <- lapply(current_terms, function(term_name) {
      update(current_model, as.formula(paste(". ~ . -", term_name)))
    })

    candidate_aic <- sapply(candidate_models, AIC)
    best_idx <- which.min(candidate_aic)
    current_model <- candidate_models[[best_idx]]

    models[[length(models) + 1]] <- current_model
    steps <- rbind(
      steps,
      data.frame(
        step_num = nrow(steps),
        step = paste("Step", nrow(steps)),
        predictors = length(attr(terms(current_model), "term.labels")),
        aic = AIC(current_model),
        removed = current_terms[best_idx],
        stringsAsFactors = FALSE
      )
    )
  }

  list(models = models, steps = steps)
}

cv_model_ladder <- function(data, response_name, ladder, k = 5, repeats = 6, seed = 4310) {
  set.seed(seed)
  cv.results <- data.frame()

  for (r in 1:repeats) {
    folds <- sample(rep(1:k, length.out = nrow(data)))

    for (j in seq_along(ladder$models)) {
      form <- formula(ladder$models[[j]])

      for (fold in 1:k) {
        train.fold <- data[folds != fold, , drop = FALSE]
        valid.fold <- data[folds == fold, , drop = FALSE]

        fit <- lm(form, data = train.fold)
        pred <- predict(fit, newdata = valid.fold)
        actual <- valid.fold[[response_name]]

        cv.results <- rbind(
          cv.results,
          data.frame(
            step = ladder$steps$step[j],
            step_num = ladder$steps$step_num[j],
            predictors = ladder$steps$predictors[j],
            repeat_id = r,
            fold_id = fold,
            RMSE = sqrt(mean((actual - pred)^2)),
            MAE = mean(abs(actual - pred))
          )
        )
      }
    }
  }

  cv.results
}

summarize_cv_results <- function(cv.results) {
  out <- aggregate(
    cbind(RMSE, MAE) ~ step + step_num + predictors,
    data = cv.results,
    FUN = mean
  )

  out$RMSE <- round(out$RMSE, 4)
  out$MAE <- round(out$MAE, 4)
  out[order(out$step_num), ]
}

bootstrap_coef_table <- function(model, data, B = 200) {
  coef.names <- names(coef(model))

  boot.mat <- replicate(B, {
    idx <- sample(1:nrow(data), replace = TRUE)
    fit <- lm(formula(model), data = data[idx, , drop = FALSE])
    coef(fit)[coef.names]
  })

  if (is.null(dim(boot.mat))) {
    boot.mat <- matrix(boot.mat, nrow = length(coef.names))
  }

  out <- data.frame(
    term = coef.names,
    estimate = coef(model),
    lower = apply(boot.mat, 1, quantile, probs = 0.025, na.rm = TRUE),
    upper = apply(boot.mat, 1, quantile, probs = 0.975, na.rm = TRUE)
  )

  out[, 2:4] <- round(out[, 2:4], 4)
  rownames(out) <- NULL
  out
}

rolling_market_stat <- function(stock_series, market_series, width = 60,
                                stat = c("beta", "correlation")) {
  stat <- match.arg(stat)

  zoo::rollapplyr(
    merge(stock = stock_series, market = market_series),
    width = width,
    by.column = FALSE,
    fill = NA,
    FUN = function(z) {
      z <- as.matrix(z)

      if (stat == "beta") {
        cov(z[, 1], z[, 2]) / var(z[, 2])
      } else {
        cor(z[, 1], z[, 2])
      }
    }
  )
}

data("HousePrices", package = "AER")
house.data <- HousePrices
index <- sample(1:2, nrow(house.data), prob = c(.8, .2), replace = TRUE)
house.train <- house.data[index == 1, ]
house.test <- house.data[index == 2, ]
house.model <- lm(price ~ ., data = house.train)
house.ladder <- build_backward_ladder(house.model)
house.cv <- cv_model_ladder(
  data = house.train,
  response_name = "price",
  ladder = house.ladder,
  k = 5,
  repeats = 6,
  seed = 4310
)
house.cv.summary <- summarize_cv_results(house.cv)
house.aic.best.idx <- which.min(house.ladder$steps$aic)
house.cv.best.idx <- which.min(house.cv.summary$RMSE)
house.model.aic.best <- house.ladder$models[[house.aic.best.idx]]
house.model.cv.best <- house.ladder$models[[house.cv.best.idx]]
house.pred.aic.best <- predict(house.model.aic.best, newdata = house.test)
house.pred.cv.best <- predict(house.model.cv.best, newdata = house.test)
house.aic.cv.rmse <- house.cv.summary$RMSE[
  match(house.ladder$steps$step[house.aic.best.idx], house.cv.summary$step)
]
house.compare <- data.frame(
  Model = c("AIC-optimal", "CV-optimal"),
  Step = c(
    house.ladder$steps$step[house.aic.best.idx],
    house.cv.summary$step[house.cv.best.idx]
  ),
  Predictors = c(
    house.ladder$steps$predictors[house.aic.best.idx],
    house.cv.summary$predictors[house.cv.best.idx]
  ),
  CV_RMSE = c(
    house.cv.summary$RMSE[match(house.ladder$steps$step[house.aic.best.idx], house.cv.summary$step)],
    house.cv.summary$RMSE[house.cv.best.idx]
  ),
  Test_RMSE = c(
    rmse(house.pred.aic.best, house.test$price),
    rmse(house.pred.cv.best, house.test$price)
  )
)
house.compare[, 4:5] <- round(house.compare[, 4:5], 4)
house.cv.labels <- paste0("S", house.cv.summary$step_num)
house.model.red <- stepAIC(house.model, direction = "both", trace = FALSE)
house.cook <- cooks.distance(house.model.red)
house.cutoff <- 4 / nrow(house.train)
house.train.new <- house.train[house.cook < house.cutoff, ]
house.model.final <- update(house.model.red, data = house.train.new)
house.prediction <- predict(house.model.final, newdata = house.test)
house.pred.before <- predict(house.model.red, newdata = house.test)
house.pred.after <- predict(house.model.final, newdata = house.test)
house.clean.model <- lm(price ~ ., data = house.train.new)
house.clean.ladder <- build_backward_ladder(house.clean.model)
house.clean.cv <- cv_model_ladder(
  data = house.train.new,
  response_name = "price",
  ladder = house.clean.ladder,
  k = 5,
  repeats = 6,
  seed = 4310
)
house.clean.cv.summary <- summarize_cv_results(house.clean.cv)
house.clean.aic.best.idx <- which.min(house.clean.ladder$steps$aic)
house.clean.cv.best.idx <- which.min(house.clean.cv.summary$RMSE)
house.clean.aic.cv.rmse <- house.clean.cv.summary$RMSE[
  match(house.clean.ladder$steps$step[house.clean.aic.best.idx], house.clean.cv.summary$step)
]
house.clean.cv.labels <- paste0("S", house.clean.cv.summary$step_num)
house.cv.ylim <- range(c(house.cv$RMSE, house.clean.cv$RMSE))
house.clean.model.aic.best <- house.clean.ladder$models[[house.clean.aic.best.idx]]
house.clean.model.cv.best <- house.clean.ladder$models[[house.clean.cv.best.idx]]
house.clean.pred.aic.best <- predict(house.clean.model.aic.best, newdata = house.test)
house.clean.pred.cv.best <- predict(house.clean.model.cv.best, newdata = house.test)
house.sensitivity <- data.frame(
  Version = c("Original training data", "Cook-filtered training data"),
  AIC_Best_Step = c(
    house.ladder$steps$step[house.aic.best.idx],
    house.clean.ladder$steps$step[house.clean.aic.best.idx]
  ),
  AIC_CV_RMSE = c(
    house.aic.cv.rmse,
    house.clean.aic.cv.rmse
  ),
  CV_Best_Step = c(
    house.cv.summary$step[house.cv.best.idx],
    house.clean.cv.summary$step[house.clean.cv.best.idx]
  ),
  CV_RMSE = c(
    house.cv.summary$RMSE[house.cv.best.idx],
    house.clean.cv.summary$RMSE[house.clean.cv.best.idx]
  )
)
house.sensitivity[, c("AIC_CV_RMSE", "CV_RMSE")] <-
  round(house.sensitivity[, c("AIC_CV_RMSE", "CV_RMSE")], 2)
house.compare.display <- house.compare[, c("Model", "Step", "Predictors", "CV_RMSE")]
names(house.compare.display) <- c("Model", "Best Step", "Predictors", "CV RMSE")
house.sensitivity.display <- house.sensitivity
names(house.sensitivity.display) <- c(
  "Training Data Used", "AIC Best Step", "AIC RMSE",
  "CV Best Step", "CV RMSE"
)
house.test.display <- data.frame(
  Version = c("Original training data", "Cook-filtered training data"),
  AIC_Best_Step = c(
    house.ladder$steps$step[house.aic.best.idx],
    house.clean.ladder$steps$step[house.clean.aic.best.idx]
  ),
  AIC_Test_RMSE = c(
    rmse(house.pred.aic.best, house.test$price),
    rmse(house.clean.pred.aic.best, house.test$price)
  ),
  CV_Best_Step = c(
    house.cv.summary$step[house.cv.best.idx],
    house.clean.cv.summary$step[house.clean.cv.best.idx]
  ),
  CV_Test_RMSE = c(
    rmse(house.pred.cv.best, house.test$price),
    rmse(house.clean.pred.cv.best, house.test$price)
  )
)
house.test.display <- data.frame(
  Version = house.test.display$Version,
  AIC_Best_Step = house.test.display$AIC_Best_Step,
  CV_Best_Step = house.test.display$CV_Best_Step,
  Held_Out_Test_RMSE = c(
    format_test_rmse(
      house.test.display$AIC_Best_Step[1], house.test.display$CV_Best_Step[1],
      house.test.display$AIC_Test_RMSE[1], house.test.display$CV_Test_RMSE[1]
    ),
    format_test_rmse(
      house.test.display$AIC_Best_Step[2], house.test.display$CV_Best_Step[2],
      house.test.display$AIC_Test_RMSE[2], house.test.display$CV_Test_RMSE[2]
    )
  )
)
names(house.test.display) <- c(
  "Training Data Used", "AIC Best Step", "CV Best Step", "Held-out Test RMSE"
)
house.cv.gap <- round(house.compare$CV_RMSE[1] - house.compare$CV_RMSE[2], 2)
house.test.flip <- round(house.compare$Test_RMSE[2] - house.compare$Test_RMSE[1], 2)
house.boot <- bootstrap_coef_table(house.model.final, house.train.new, B = 200)

house.summary <- data.frame(
  Metric = c("Adjusted R2", "Residual s", "Outliers Removed", "Test RMSE", "Test R2"),
  Value = c(
    round(summary(house.model.final)$adj.r.squared, 4),
    round(summary(house.model.final)$sigma, 2),
    length(which(house.cook > house.cutoff)),
    round(rmse(house.prediction, house.test$price), 2),
    round(1 - sum((house.test$price - house.prediction)^2) /
      sum((house.test$price - mean(house.test$price))^2), 4)
  )
)

house.key <- get_term_table(house.model.final)
house.diag <- get_diag(house.model.final)
house.predictors <- paste(attr(terms(house.model.final), "term.labels"), collapse = ", ")
house.removed <- setdiff(attr(terms(house.model), "term.labels"), attr(terms(house.model.final), "term.labels"))
house.removed.text <- if (length(house.removed) == 0) "None from the full model" else paste(house.removed, collapse = ", ")

house.new <- data.frame(
  lotsize = c(5000, 8000),
  bedrooms = c(3, 4),
  bathrooms = c(1, 2),
  stories = c(1, 2),
  driveway = factor(c("yes", "yes"), levels = levels(house.data$driveway)),
  recreation = factor(c("no", "yes"), levels = levels(house.data$recreation)),
  fullbase = factor(c("no", "yes"), levels = levels(house.data$fullbase)),
  gasheat = factor(c("no", "yes"), levels = levels(house.data$gasheat)),
  aircon = factor(c("no", "yes"), levels = levels(house.data$aircon)),
  garage = c(0, 2),
  prefer = factor(c("no", "yes"), levels = levels(house.data$prefer))
)

house.conf <- predict(house.model.final, newdata = house.new, interval = "confidence")
house.pred <- predict(house.model.final, newdata = house.new, interval = "prediction")

house.pred.table <- data.frame(
  Scenario = c("Smaller home", "Larger home"),
  Fit = round(house.conf[, "fit"], 0),
  CI = paste0("(", round(house.conf[, "lwr"], 0), ", ", round(house.conf[, "upr"], 0), ")"),
  PI = paste0("(", round(house.pred[, "lwr"], 0), ", ", round(house.pred[, "upr"], 0), ")")
)

house.actual.conf <- predict(house.model.final, interval = "confidence")
house.actual.pred <- predict(house.model.final, interval = "prediction")
house.order <- order(house.actual.conf[, "fit"])
house.fit.x <- house.actual.conf[house.order, "fit"]
house.new.fit <- house.conf[, "fit"]
house.process <- data.frame(
  Step = c(
    "Split data",
    "Build full model",
    "Build model ladder",
    "Run repeated CV",
    "Compare AIC and CV",
    "Remove outliers",
    "Check residuals",
    "Bootstrap intervals",
    "Predict new homes"
  ),
  Detail = c(
    paste(nrow(house.train), "train /", nrow(house.test), "test"),
    paste(length(attr(terms(house.model), "term.labels")), "starting predictors"),
    paste(nrow(house.ladder$steps), "nested steps"),
    "5 folds x 6 repeats",
    paste(house.compare$Step[1], "vs", house.compare$Step[2]),
    paste(length(which(house.cook > house.cutoff)), "points above Cook cutoff"),
    "QQ, BP, and DW tests",
    "200 bootstrap resamples",
    "2 cases with CI and PI"
  )
)

data("CASchools", package = "AER")
schools.data <- CASchools
schools.data$score <- (schools.data$read + schools.data$math) / 2
schools.data$str <- schools.data$students / schools.data$teachers
schools.data$computer.rate <- schools.data$computer / schools.data$students
schools.data <- schools.data[, c(
  "score", "str", "calworks", "lunch", "computer.rate",
  "expenditure", "income", "english", "grades"
)]

index <- sample(1:2, nrow(schools.data), prob = c(.8, .2), replace = TRUE)
schools.train <- schools.data[index == 1, ]
schools.test <- schools.data[index == 2, ]
schools.model <- lm(score ~ ., data = schools.train)
schools.ladder <- build_backward_ladder(schools.model)
schools.cv <- cv_model_ladder(
  data = schools.train,
  response_name = "score",
  ladder = schools.ladder,
  k = 5,
  repeats = 6,
  seed = 4310
)
schools.cv.summary <- summarize_cv_results(schools.cv)
schools.aic.best.idx <- which.min(schools.ladder$steps$aic)
schools.cv.best.idx <- which.min(schools.cv.summary$RMSE)
schools.model.aic.best <- schools.ladder$models[[schools.aic.best.idx]]
schools.model.cv.best <- schools.ladder$models[[schools.cv.best.idx]]
schools.pred.aic.best <- predict(schools.model.aic.best, newdata = schools.test)
schools.pred.cv.best <- predict(schools.model.cv.best, newdata = schools.test)
schools.aic.cv.rmse <- schools.cv.summary$RMSE[
  match(schools.ladder$steps$step[schools.aic.best.idx], schools.cv.summary$step)
]
schools.compare <- data.frame(
  Model = c("AIC-optimal", "CV-optimal"),
  Step = c(
    schools.ladder$steps$step[schools.aic.best.idx],
    schools.cv.summary$step[schools.cv.best.idx]
  ),
  Predictors = c(
    schools.ladder$steps$predictors[schools.aic.best.idx],
    schools.cv.summary$predictors[schools.cv.best.idx]
  ),
  CV_RMSE = c(
    schools.cv.summary$RMSE[match(schools.ladder$steps$step[schools.aic.best.idx], schools.cv.summary$step)],
    schools.cv.summary$RMSE[schools.cv.best.idx]
  ),
  Test_RMSE = c(
    rmse(schools.pred.aic.best, schools.test$score),
    rmse(schools.pred.cv.best, schools.test$score)
  )
)
schools.compare[, 4:5] <- round(schools.compare[, 4:5], 4)
schools.cv.labels <- paste0("S", schools.cv.summary$step_num)
schools.model.red <- stepAIC(schools.model, direction = "both", trace = FALSE)
schools.cook <- cooks.distance(schools.model.red)
schools.cutoff <- 4 / nrow(schools.train)
schools.train.new <- schools.train[schools.cook < schools.cutoff, ]
schools.model.final <- update(schools.model.red, data = schools.train.new)
schools.prediction <- predict(schools.model.final, newdata = schools.test)
schools.pred.before <- predict(schools.model.red, newdata = schools.test)
schools.pred.after <- predict(schools.model.final, newdata = schools.test)
schools.clean.model <- lm(score ~ ., data = schools.train.new)
schools.clean.ladder <- build_backward_ladder(schools.clean.model)
schools.clean.cv <- cv_model_ladder(
  data = schools.train.new,
  response_name = "score",
  ladder = schools.clean.ladder,
  k = 5,
  repeats = 6,
  seed = 4310
)
schools.clean.cv.summary <- summarize_cv_results(schools.clean.cv)
schools.clean.aic.best.idx <- which.min(schools.clean.ladder$steps$aic)
schools.clean.cv.best.idx <- which.min(schools.clean.cv.summary$RMSE)
schools.clean.aic.cv.rmse <- schools.clean.cv.summary$RMSE[
  match(schools.clean.ladder$steps$step[schools.clean.aic.best.idx], schools.clean.cv.summary$step)
]
schools.clean.cv.labels <- paste0("S", schools.clean.cv.summary$step_num)
schools.cv.ylim <- range(c(schools.cv$RMSE, schools.clean.cv$RMSE))
schools.clean.model.aic.best <- schools.clean.ladder$models[[schools.clean.aic.best.idx]]
schools.clean.model.cv.best <- schools.clean.ladder$models[[schools.clean.cv.best.idx]]
schools.clean.pred.aic.best <- predict(schools.clean.model.aic.best, newdata = schools.test)
schools.clean.pred.cv.best <- predict(schools.clean.model.cv.best, newdata = schools.test)
schools.sensitivity <- data.frame(
  Version = c("Original training data", "Cook-filtered training data"),
  AIC_Best_Step = c(
    schools.ladder$steps$step[schools.aic.best.idx],
    schools.clean.ladder$steps$step[schools.clean.aic.best.idx]
  ),
  AIC_CV_RMSE = c(
    schools.aic.cv.rmse,
    schools.clean.aic.cv.rmse
  ),
  CV_Best_Step = c(
    schools.cv.summary$step[schools.cv.best.idx],
    schools.clean.cv.summary$step[schools.clean.cv.best.idx]
  ),
  CV_RMSE = c(
    schools.cv.summary$RMSE[schools.cv.best.idx],
    schools.clean.cv.summary$RMSE[schools.clean.cv.best.idx]
  )
)
schools.sensitivity[, c("AIC_CV_RMSE", "CV_RMSE")] <-
  round(schools.sensitivity[, c("AIC_CV_RMSE", "CV_RMSE")], 2)
schools.compare.display <- schools.compare[, c("Model", "Step", "Predictors", "CV_RMSE")]
names(schools.compare.display) <- c("Model", "Best Step", "Predictors", "CV RMSE")
schools.sensitivity.display <- schools.sensitivity
names(schools.sensitivity.display) <- c(
  "Training Data Used", "AIC Best Step", "AIC RMSE",
  "CV Best Step", "CV RMSE"
)
schools.test.display <- data.frame(
  Version = c("Original training data", "Cook-filtered training data"),
  AIC_Best_Step = c(
    schools.ladder$steps$step[schools.aic.best.idx],
    schools.clean.ladder$steps$step[schools.clean.aic.best.idx]
  ),
  AIC_Test_RMSE = c(
    rmse(schools.pred.aic.best, schools.test$score),
    rmse(schools.clean.pred.aic.best, schools.test$score)
  ),
  CV_Best_Step = c(
    schools.cv.summary$step[schools.cv.best.idx],
    schools.clean.cv.summary$step[schools.clean.cv.best.idx]
  ),
  CV_Test_RMSE = c(
    rmse(schools.pred.cv.best, schools.test$score),
    rmse(schools.clean.pred.cv.best, schools.test$score)
  )
)
schools.test.display <- data.frame(
  Version = schools.test.display$Version,
  AIC_Best_Step = schools.test.display$AIC_Best_Step,
  CV_Best_Step = schools.test.display$CV_Best_Step,
  Held_Out_Test_RMSE = c(
    format_test_rmse(
      schools.test.display$AIC_Best_Step[1], schools.test.display$CV_Best_Step[1],
      schools.test.display$AIC_Test_RMSE[1], schools.test.display$CV_Test_RMSE[1]
    ),
    format_test_rmse(
      schools.test.display$AIC_Best_Step[2], schools.test.display$CV_Best_Step[2],
      schools.test.display$AIC_Test_RMSE[2], schools.test.display$CV_Test_RMSE[2]
    )
  )
)
names(schools.test.display) <- c(
  "Training Data Used", "AIC Best Step", "CV Best Step", "Held-out Test RMSE"
)
schools.boot <- bootstrap_coef_table(schools.model.final, schools.train.new, B = 200)

schools.summary <- data.frame(
  Metric = c("Adjusted R2", "Residual s", "Outliers Removed", "Test RMSE", "Test R2"),
  Value = c(
    round(summary(schools.model.final)$adj.r.squared, 4),
    round(summary(schools.model.final)$sigma, 2),
    length(which(schools.cook > schools.cutoff)),
    round(rmse(schools.prediction, schools.test$score), 2),
    round(1 - sum((schools.test$score - schools.prediction)^2) /
      sum((schools.test$score - mean(schools.test$score))^2), 4)
  )
)

schools.key <- get_term_table(schools.model.final)
schools.diag <- get_diag(schools.model.final)
schools.predictors <- paste(attr(terms(schools.model.final), "term.labels"), collapse = ", ")
schools.removed <- setdiff(attr(terms(schools.model), "term.labels"), attr(terms(schools.model.final), "term.labels"))
schools.removed.text <- if (length(schools.removed) == 0) "None from the full model" else paste(schools.removed, collapse = ", ")

schools.shared.terms <- c("str", "calworks", "lunch", "income", "english")
schools.shared.formula <- as.formula(
  paste("score ~", paste(schools.shared.terms, collapse = " + "))
)
schools.strata <- levels(schools.train.new$grades)
schools.stratified.fits <- lapply(schools.strata, function(g) {
  lm(schools.shared.formula, data = subset(schools.train.new, grades == g))
})
names(schools.stratified.fits) <- schools.strata

schools.stratified.summary <- do.call(
  rbind,
  lapply(schools.strata, function(g) {
    sub <- subset(schools.train.new, grades == g)
    fit <- schools.stratified.fits[[g]]

    data.frame(
      District.Type = g,
      N = nrow(sub),
      Mean.Score = mean(sub$score),
      Adjusted.R2 = summary(fit)$adj.r.squared,
      Residual.s = summary(fit)$sigma
    )
  })
)
schools.stratified.summary$Mean.Score <- round(schools.stratified.summary$Mean.Score, 2)
schools.stratified.summary$Adjusted.R2 <- round(schools.stratified.summary$Adjusted.R2, 4)
schools.stratified.summary$Residual.s <- round(schools.stratified.summary$Residual.s, 2)
schools.stratified.summary.display <- schools.stratified.summary
names(schools.stratified.summary.display) <- c(
  "District Type", "N", "Mean Score", "Adjusted R2", "Residual s"
)

schools.within_group.compare <- do.call(
  rbind,
  lapply(schools.strata, function(g) {
    sub <- subset(schools.train.new, grades == g)
    fit <- schools.stratified.fits[[g]]
    pooled.pred <- predict(schools.model.final, newdata = sub)
    strat.pred <- predict(fit, newdata = sub)

    data.frame(
      District.Type = g,
      N = nrow(sub),
      Pooled.Within.R2 = subgroup_r2(sub$score, pooled.pred),
      Stratified.Within.R2 = subgroup_r2(sub$score, strat.pred),
      Gain = subgroup_r2(sub$score, strat.pred) - subgroup_r2(sub$score, pooled.pred)
    )
  })
)
schools.within_group.compare[, 3:5] <- round(schools.within_group.compare[, 3:5], 4)
schools.within_group.compare.display <- schools.within_group.compare
names(schools.within_group.compare.display) <- c(
  "District Type", "N", "Pooled Within-Group R2",
  "Stratified Within-Group R2", "R2 Gain from Splitting"
)
kk06.gain <- schools.within_group.compare$Gain[schools.within_group.compare$District.Type == "KK-06"]
kk08.gain <- schools.within_group.compare$Gain[schools.within_group.compare$District.Type == "KK-08"]

schools.stratified.term.display <- data.frame(Term = schools.shared.terms)
for (g in schools.strata) {
  fit.summary <- coef(summary(schools.stratified.fits[[g]]))[schools.shared.terms, , drop = FALSE]
  schools.stratified.term.display[[paste0(g, " Estimate")]] <- sprintf(
    "%.4f",
    fit.summary[, "Estimate"]
  )
  schools.stratified.term.display[[paste0(g, " P Value")]] <- trimws(format.pval(
    fit.summary[, "Pr(>|t|)"],
    digits = 3,
    eps = 1e-16
  ))
}
names(schools.stratified.term.display) <- c(
  "Term", "KK-06 Estimate", "KK-06 P Value", "KK-08 Estimate", "KK-08 P Value"
)
schools.stratified.term.compact <- data.frame(
  Term = schools.stratified.term.display$Term,
  `KK-06 Est (p)` = paste0(
    schools.stratified.term.display$`KK-06 Estimate`,
    " (", schools.stratified.term.display$`KK-06 P Value`, ")"
  ),
  `KK-08 Est (p)` = paste0(
    schools.stratified.term.display$`KK-08 Estimate`,
    " (", schools.stratified.term.display$`KK-08 P Value`, ")"
  ),
  check.names = FALSE
)

schools.stratified.coef <- do.call(
  rbind,
  lapply(schools.strata, function(g) {
    fit <- schools.stratified.fits[[g]]
    ci <- confint(fit)

    data.frame(
      District.Type = g,
      Term = schools.shared.terms,
      Estimate = coef(fit)[schools.shared.terms],
      Lower = ci[schools.shared.terms, 1],
      Upper = ci[schools.shared.terms, 2]
    )
  })
)
schools.stratified.coef$Term <- factor(
  schools.stratified.coef$Term,
  levels = rev(schools.shared.terms)
)
schools.stratified.coef$District.Type <- factor(
  schools.stratified.coef$District.Type,
  levels = schools.strata
)

schools.interaction.model <- lm(
  score ~ (str + calworks + lunch + income + english) * grades,
  data = schools.train.new
)
schools.interaction.anova <- anova(schools.model.final, schools.interaction.model)
schools.interaction.anova.display <- data.frame(
  Model = c("Main-effects pooled model", "Add 5 district-type interactions"),
  Residual.DF = schools.interaction.anova$Res.Df,
  RSS = round(schools.interaction.anova$RSS, 2),
  F = c("", sprintf("%.4f", schools.interaction.anova$F[2])),
  P.Value = c(
    "",
    trimws(format.pval(schools.interaction.anova$`Pr(>F)`[2], digits = 3, eps = 1e-16))
  )
)

schools.interaction.coef <- coef(summary(schools.interaction.model))
schools.interaction.rows <- grep(":", rownames(schools.interaction.coef))
schools.interaction.terms.display <- data.frame(
  Term = rownames(schools.interaction.coef)[schools.interaction.rows],
  Estimate = sprintf("%.4f", schools.interaction.coef[schools.interaction.rows, "Estimate"]),
  P.Value = trimws(format.pval(
    schools.interaction.coef[schools.interaction.rows, "Pr(>|t|)"],
    digits = 3,
    eps = 1e-16
  ))
)
schools.interaction.p.value <- trimws(format.pval(
  schools.interaction.anova$`Pr(>F)`[2],
  digits = 3,
  eps = 1e-16
))

schools.interaction.slope.display <- do.call(
  rbind,
  lapply(schools.shared.terms, function(term) {
    delta.name <- paste0(term, ":gradesKK-08")
    kk06 <- linear_combo_stats(
      schools.interaction.model,
      setNames(1, term)
    )
    kk08 <- linear_combo_stats(
      schools.interaction.model,
      setNames(c(1, 1), c(term, delta.name))
    )

    data.frame(
      Term = term,
      KK.06.Slope = sprintf("%.4f", kk06["Estimate"]),
      KK.06.P.Value = trimws(format.pval(kk06["P.Value"], digits = 3, eps = 1e-16)),
      KK.08.Slope = sprintf("%.4f", kk08["Estimate"]),
      KK.08.P.Value = trimws(format.pval(kk08["P.Value"], digits = 3, eps = 1e-16)),
      Delta.KK.08.minus.KK.06 = sprintf("%.4f", schools.interaction.coef[delta.name, "Estimate"]),
      Delta.P.Value = trimws(format.pval(
        schools.interaction.coef[delta.name, "Pr(>|t|)"],
        digits = 3,
        eps = 1e-16
      ))
    )
  })
)
names(schools.interaction.slope.display) <- c(
  "Term", "KK-06 Slope", "KK-06 P Value", "KK-08 Slope",
  "KK-08 P Value", "Delta (08-06)", "Delta P Value"
)
schools.interaction.slope.compact <- data.frame(
  Term = schools.interaction.slope.display$Term,
  `KK-06 Slope (p)` = paste0(
    schools.interaction.slope.display$`KK-06 Slope`,
    " (", schools.interaction.slope.display$`KK-06 P Value`, ")"
  ),
  `KK-08 Slope (p)` = paste0(
    schools.interaction.slope.display$`KK-08 Slope`,
    " (", schools.interaction.slope.display$`KK-08 P Value`, ")"
  ),
  `Delta 08-06 (p)` = paste0(
    schools.interaction.slope.display$`Delta (08-06)`,
    " (", schools.interaction.slope.display$`Delta P Value`, ")"
  ),
  check.names = FALSE
)

schools.interaction.grid <- expand.grid(
  income = seq(min(schools.train.new$income), max(schools.train.new$income), length.out = 100),
  grades = factor(schools.strata, levels = schools.strata)
)
schools.interaction.grid$str <- median(schools.train.new$str)
schools.interaction.grid$calworks <- median(schools.train.new$calworks)
schools.interaction.grid$lunch <- median(schools.train.new$lunch)
schools.interaction.grid$english <- median(schools.train.new$english)
schools.interaction.grid$fit <- predict(
  schools.interaction.model,
  newdata = schools.interaction.grid
)

schools.new <- data.frame(
  str = c(18, 22),
  calworks = c(10, 45),
  lunch = c(25, 70),
  computer.rate = c(0.25, 0.12),
  expenditure = c(5500, 5000),
  income = c(20, 12),
  english = c(5, 30),
  grades = factor(c("KK-06", "KK-08"), levels = levels(schools.data$grades))
)

schools.conf <- predict(
  schools.model.final,
  newdata = schools.new,
  interval = "confidence"
)

schools.pred <- predict(
  schools.model.final,
  newdata = schools.new,
  interval = "prediction"
)

schools.pred.table <- data.frame(
  Scenario = c("Lower-risk district", "Higher-risk district"),
  Fit = round(schools.conf[, "fit"], 2),
  CI = paste0("(", round(schools.conf[, "lwr"], 2), ", ", round(schools.conf[, "upr"], 2), ")"),
  PI = paste0("(", round(schools.pred[, "lwr"], 2), ", ", round(schools.pred[, "upr"], 2), ")")
)

schools.actual.conf <- predict(schools.model.final, interval = "confidence")
schools.actual.pred <- predict(schools.model.final, interval = "prediction")
schools.order <- order(schools.actual.conf[, "fit"])
schools.fit.x <- schools.actual.conf[schools.order, "fit"]
schools.new.fit <- schools.conf[, "fit"]
schools.process <- data.frame(
  Step = c(
    "Split data",
    "Build full model",
    "Build model ladder",
    "Run repeated CV",
    "Compare AIC and CV",
    "Remove outliers",
    "Check residuals",
    "Bootstrap intervals",
    "Predict districts"
  ),
  Detail = c(
    paste(nrow(schools.train), "train /", nrow(schools.test), "test"),
    paste(length(attr(terms(schools.model), "term.labels")), "starting predictors"),
    paste(nrow(schools.ladder$steps), "nested steps"),
    "5 folds x 6 repeats",
    paste(schools.compare$Step[1], "vs", schools.compare$Step[2]),
    paste(length(which(schools.cook > schools.cutoff)), "points above Cook cutoff"),
    "QQ, BP, and DW tests",
    "200 bootstrap resamples",
    "2 cases with CI and PI"
  )
)
```

```{r}
finance.stocks <- data.frame(
  Ticker = c("JPM", "JNJ", "XOM", "WMT", "CAT", "BA"),
  Company = c(
    "JPMorgan Chase",
    "Johnson & Johnson",
    "Exxon Mobil",
    "Walmart",
    "Caterpillar",
    "Boeing Company"
  ),
  Industry = c(
    "Financial Services",
    "Healthcare",
    "Energy",
    "Consumer Defensive",
    "Industrials",
    "Aerospace"
  )
)
# work
JPM.f <- getSymbols("JPM", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
JNJ.f <- getSymbols("JNJ", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
XOM.f <- getSymbols("XOM", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
WMT.f <- getSymbols("WMT", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
CAT.f <- getSymbols("CAT", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
BA.f <- getSymbols("BA", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)
GSPC.f <- getSymbols("^GSPC", src = "yahoo", from = "2023-01-01", to = "2025-12-31", auto.assign = FALSE)

finance.close <- na.omit(merge(
  JPM = Cl(JPM.f),
  JNJ = Cl(JNJ.f),
  XOM = Cl(XOM.f),
  WMT = Cl(WMT.f),
  CAT = Cl(CAT.f),
  BA = Cl(BA.f),
  SP500 = Cl(GSPC.f)
))
colnames(finance.close) <- c("JPM", "JNJ", "XOM", "WMT", "CAT", "BA", "SP500")

finance.ret <- na.omit(merge(
  dailyReturn(Cl(JPM.f), type = "arithmetic"),
  dailyReturn(Cl(JNJ.f), type = "arithmetic"),
  dailyReturn(Cl(XOM.f), type = "arithmetic"),
  dailyReturn(Cl(WMT.f), type = "arithmetic"),
  dailyReturn(Cl(CAT.f), type = "arithmetic"),
  dailyReturn(Cl(BA.f), type = "arithmetic"),
  dailyReturn(Cl(GSPC.f), type = "arithmetic")
))
colnames(finance.ret) <- c("JPM", "JNJ", "XOM", "WMT", "CAT","BA", "SP500")
finance.ret.mat <- coredata(finance.ret)

finance.annual <- data.frame(
  Ticker = colnames(finance.ret),
  Annual_Return = round(colMeans(finance.ret.mat) * 252, 4),
  Annual_Risk = round(apply(finance.ret.mat, 2, sd) * sqrt(252), 4)
)
rownames(finance.annual) <- NULL

finance.corr <- round(cor(finance.ret.mat), 3)
finance.cum <- xts(
  apply(1 + finance.ret.mat, 2, cumprod),
  order.by = index(finance.ret)
)

finance.capm <- data.frame(
  Ticker = finance.stocks$Ticker,
  Alpha_Daily = rep(NA, 6),
  Alpha_Annual = rep(NA, 6),
  Beta = rep(NA, 6),
  R_Squared = rep(NA, 6)
)

for (i in 1:6) {
  stock.name <- finance.stocks$Ticker[i]
  model <- lm(finance.ret.mat[, stock.name] ~ finance.ret.mat[, "SP500"])
  finance.capm$Alpha_Daily[i] <- coef(model)[1]
  finance.capm$Alpha_Annual[i] <- coef(model)[1] * 252
  finance.capm$Beta[i] <- coef(model)[2]
  finance.capm$R_Squared[i] <- summary(model)$r.squared
}

finance.capm[, 2:5] <- round(finance.capm[, 2:5], 4)
rownames(finance.capm) <- NULL

finance.compare <- data.frame(
  Ticker = finance.stocks$Ticker,
  Annual_Return = finance.annual$Annual_Return[match(finance.stocks$Ticker, finance.annual$Ticker)],
  Annual_Risk = finance.annual$Annual_Risk[match(finance.stocks$Ticker, finance.annual$Ticker)],
  Corr_SP500 = round(cor(finance.ret.mat[, finance.stocks$Ticker], finance.ret.mat[, "SP500"]), 4)
)
rownames(finance.compare) <- NULL
finance.label.pos <- c(2, 4, 1, 2, 4)
finance.annual.display <- finance.annual
names(finance.annual.display) <- c("Ticker", "Annual Return", "Annual Risk")
finance.compare.display <- finance.compare
names(finance.compare.display) <- c("Ticker", "Annual Return", "Annual Risk", "Corr with SP500")
finance.capm.display <- finance.capm
names(finance.capm.display) <- c("Ticker", "Alpha Daily", "Alpha Annual", "Beta", "R Squared")
finance.top.return <- finance.compare$Ticker[which.max(finance.compare$Annual_Return)]
finance.high.risk <- finance.compare$Ticker[which.max(finance.compare$Annual_Risk)]
finance.low.risk <- finance.compare$Ticker[which.min(finance.compare$Annual_Risk)]
finance.annual.label.pos <- ifelse(finance.annual$Ticker == finance.high.risk, 2, 4)
finance.high.corr <- finance.compare$Ticker[which.max(finance.compare$Corr_SP500)]
finance.low.corr <- finance.compare$Ticker[which.min(finance.compare$Corr_SP500)]
finance.high.beta <- finance.capm$Ticker[which.max(finance.capm$Beta)]
finance.high.rsq <- finance.capm$Ticker[which.max(finance.capm$R_Squared)]
finance.best.cum <- finance.stocks$Ticker[
  which.max(as.numeric(last(finance.cum[, finance.stocks$Ticker])))
]
finance.worst.cum <- finance.stocks$Ticker[
  which.min(as.numeric(last(finance.cum[, finance.stocks$Ticker])))
]

finance.process <- data.frame(
  Step = c(
    "Pull close prices",
    "Convert to daily returns",
    "Annualize return and risk",
    "Compare correlation to market",
    "Plot cumulative returns",
    "Run CAPM regressions",
    "Track rolling beta",
    "Track rolling correlation"
  ),
  Detail = c(
    "Yahoo Finance, 2023-2025",
    "Arithmetic daily returns",
    "Mean*252 and sd*sqrt(252)",
    "Corr(stock, SP500)",
    "Growth of $1 over time",
    "alpha, beta, and R2",
    "60-day beta windows",
    "60-day correlation windows"
  )
)

finance.roll.beta <- na.omit(merge(
  JPM = rolling_market_stat(finance.ret[, "JPM"], finance.ret[, "SP500"], stat = "beta"),
  JNJ = rolling_market_stat(finance.ret[, "JNJ"], finance.ret[, "SP500"], stat = "beta"),
  XOM = rolling_market_stat(finance.ret[, "XOM"], finance.ret[, "SP500"], stat = "beta"),
  WMT = rolling_market_stat(finance.ret[, "WMT"], finance.ret[, "SP500"], stat = "beta"),
  CAT = rolling_market_stat(finance.ret[, "CAT"], finance.ret[, "SP500"], stat = "beta"),
  BA = rolling_market_stat(finance.ret[, "BA"], finance.ret[, "SP500"], stat = "beta")
))

finance.roll.corr <- na.omit(merge(
  JPM = rolling_market_stat(finance.ret[, "JPM"], finance.ret[, "SP500"], stat = "correlation"),
  JNJ = rolling_market_stat(finance.ret[, "JNJ"], finance.ret[, "SP500"], stat = "correlation"),
  XOM = rolling_market_stat(finance.ret[, "XOM"], finance.ret[, "SP500"], stat = "correlation"),
  WMT = rolling_market_stat(finance.ret[, "WMT"], finance.ret[, "SP500"], stat = "correlation"),
  CAT = rolling_market_stat(finance.ret[, "CAT"], finance.ret[, "SP500"], stat = "correlation"),
  BA = rolling_market_stat(finance.ret[, "BA"], finance.ret[, "SP500"], stat = "correlation")
))
```

class: center, middle, inverse

# STAT 4310 Regression Project

## Part I: Two Regression Datasets
## Part II: Finance Application

<span class="accent">HousePrices, CASchools, and a 5-stock NYSE portfolio</span>

---

## Roadmap

- Part I uses two medium-sized datasets with optimal-model workflows.
- Part II uses five NYSE stocks from different industries plus the
  S&P 500 benchmark.
- Every section shows methods, interim process steps, diagnostics,
  extra validation, and final interpretation.

---

class: small-table compact

## Part I: Regression Methods

```{r}
kable(
  data.frame(
    Step = c(
      "1. Describe data",
      "2. Split Train/Test",
      "3. Build full model",
      "4. Build CV ladder",
      "5. Compare AIC vs CV",
      "6. Remove outliers",
      "7. Check sensitivity",
      "8. Check residuals",
      "9. Bootstrap intervals",
      "10. Predict new cases"
    ),
    Method = c(
      "response, predictors, missingness",
      "80% / 20%",
      "lm()",
      "nested backward models + repeated CV",
      "AIC-optimal vs CV-optimal",
      "Cook's distance",
      "before vs after Cook filtering",
      "QQ, BP, and DW tests",
      "bootstrap coefficient CIs",
      "confidence and prediction intervals"
    )
  )
)
```

---

## HousePrices: Dataset Snapshot

.pull-left[
- `546` observations and `12` variables
- Response: `price`
- `lotsize`: lot size of the house
- `bedrooms`, `bathrooms`, `stories`, `garage`: size/layout features
- `driveway`, `recreation`, `fullbase`, `gasheat`, `aircon`, `prefer`:
  yes/no housing amenities or location preference indicators
- Missing values: `0`
- The mix of structural and amenity variables makes the dataset useful
  for both interpretation and prediction.
]

.pull-right[
```{r}
kable(
  data.frame(
    Item = c("Rows", "Columns", "Missing"),
    Value = c(nrow(house.data), ncol(house.data), sum(is.na(house.data)))
  ),
  row.names = FALSE
)
```
]

---

class: plot-slide

## HousePrices: Exploratory Overview

```{r fig.width=9.5, fig.height=4.8}
par(mfrow = c(1, 3), mar = c(4.5, 4, 3, 1))

hist(
  house.data$price,
  breaks = 30,
  col = "#2a6f97",
  border = "white",
  xlab = "Price",
  main = "Distribution of Price"
)

plot(
  house.data$lotsize,
  house.data$price,
  pch = 16,
  col = adjustcolor("#123c55", alpha.f = 0.5),
  xlab = "Lot Size",
  ylab = "Price",
  main = "Price vs Lot Size"
)

plot(
  jitter(house.data$bedrooms),
  house.data$price,
  pch = 16,
  col = adjustcolor("#b83227", alpha.f = 0.5),
  xlab = "Bedrooms",
  ylab = "Price",
  main = "Price vs Bedrooms"
)

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- Price is right-skewed: most homes are modestly priced with a long tail of high-value sales.
- Lot size and price have a positive association, but with wide spread showing other features matter too.
- Bedroom count alone is a weak predictor — amenity combinations are more informative.

---

class: small-table

## HousePrices: Interim Process

```{r}
kable(house.process, row.names = FALSE)
```

- The ordering matters: choose the model on training data, check
  influential cases, then touch the held-out test set only once.

---

class: plot-slide

## HousePrices: CV Across the Selection Ladder

```{r fig.width=9.2, fig.height=4.7}
par(mar = c(7, 4.5, 4, 1))

boxplot(
  RMSE ~ factor(step_num, labels = house.cv.labels),
  data = house.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  main = "Repeated 5-fold CV on the training-set model ladder"
)

lines(seq_along(house.cv.labels), house.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_along(house.cv.labels), house.cv.summary$RMSE, pch = 16, col = "#123c55")
points(house.cv.best.idx, house.cv.summary$RMSE[house.cv.best.idx], pch = 19, col = "#0b6e4f", cex = 1.2)

abline(h = house.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "CV-best step", "AIC-selected step mean CV RMSE"),
  col = c("#123c55", "#0b6e4f", "#b83227"),
  pch = c(16, 19, NA),
  lty = c(1, NA, 2),
  lwd = c(2, NA, 2),
  bty = "n"
)

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

- `S0` is the full model.
- Each later step removes one more predictor from a nested ladder.
- Each step shows `30` validation RMSE values because we run `5` folds
  and repeat that `6` times.
- The red dashed line marks the AIC-selected step's mean CV RMSE.
- The green point marks the true CV minimum, which is `S1` here.
- One early predictor removal improves validation RMSE, but additional
  trimming starts to give up signal.

---

class: small-table

## HousePrices: AIC Step vs CV Step

```{r}
kable(house.compare.display, row.names = FALSE)
```

- AIC and CV are compared on the same training-set ladder.
- The held-out 20% test set is used only for the final check.
- For `HousePrices`, the CV gap is only `r house.cv.gap`, so the test
  flip of `r house.test.flip` is normal one-split noise rather than a
  contradiction.
- In practice, both criteria are pointing to nearly the same part of the
  ladder rather than arguing for very different housing models.

---

class: small-table

## HousePrices: Sensitivity to Cook Filtering

```{r}
kable(house.sensitivity.display, row.names = FALSE)
```

- The upper table asks whether influential sales change model choice or
  mostly change the size of the validation error.

```{r}
kable(house.test.display, row.names = FALSE)
```

- In the upper table, both RMSE columns are repeated-CV validation RMSE.
- In the lower table, the RMSE is computed on the held-out 20% test set
  by comparing predictions to the actual observed response values.
- When AIC and CV pick the same step, that test RMSE is shown once.
- After Cook filtering, the validation error drops without changing the
  overall selection story.

---

class: small-table plot-slide

## HousePrices: CV Ladder Before and After Cook Filtering

```{r fig.width=9.2, fig.height=4.6}
par(mfrow = c(1, 3), mar = c(7, 4.2, 3.5, 1))

boxplot(
  RMSE ~ factor(step_num, labels = house.cv.labels),
  data = house.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  ylim = house.cv.ylim,
  main = "Before Cook filtering"
)
lines(seq_len(nrow(house.cv.summary)), house.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_len(nrow(house.cv.summary)), house.cv.summary$RMSE, pch = 16, col = "#123c55")
abline(h = house.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "AIC-selected step"),
  col = c("#123c55", "#b83227"),
  pch = c(16, NA),
  lty = c(1, 2),
  lwd = c(2, 2),
  bty = "n",
  cex = 0.8
)

plot(
  house.cv.summary$step_num,
  house.cv.summary$RMSE,
  type = "b",
  pch = 16,
  lwd = 2,
  col = "gray40",
  xaxt = "n",
  xlab = "",
  ylab = "Mean validation RMSE",
  ylim = house.cv.ylim,
  main = "Mean CV RMSE overlay"
)
lines(
  house.clean.cv.summary$step_num,
  house.clean.cv.summary$RMSE,
  type = "b",
  pch = 17,
  lwd = 2,
  col = "#0b6e4f"
)
abline(h = house.aic.cv.rmse, col = "gray40", lty = 2, lwd = 2)
abline(h = house.clean.aic.cv.rmse, col = "#0b6e4f", lty = 2, lwd = 2)
axis(1, at = house.cv.summary$step_num, labels = house.cv.labels, las = 2, cex.axis = 0.8)
legend(
  "topleft",
  legend = c("Before Cook", "After Cook", "Before AIC RMSE", "After AIC RMSE"),
  col = c("gray40", "#0b6e4f", "gray40", "#0b6e4f"),
  pch = c(16, 17, NA, NA),
  lty = c(1, 1, 2, 2),
  lwd = c(2, 2, 2, 2),
  bty = "n",
  cex = 0.8
)

boxplot(
  RMSE ~ factor(step_num, labels = house.clean.cv.labels),
  data = house.clean.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  ylim = house.cv.ylim,
  main = "After Cook filtering"
)
lines(seq_len(nrow(house.clean.cv.summary)), house.clean.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_len(nrow(house.clean.cv.summary)), house.clean.cv.summary$RMSE, pch = 16, col = "#123c55")
abline(h = house.clean.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "AIC-selected step"),
  col = c("#123c55", "#b83227"),
  pch = c(16, NA),
  lty = c(1, 2),
  lwd = c(2, 2),
  bty = "n",
  cex = 0.8
)

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- The whole ladder shifts down after filtering, so the influential cases
  were hurting validation performance across multiple candidate models.
- Because the whole curve moves down, Cook filtering improves model
  stability rather than just rescuing one lucky final step.

---

class: small-table compact

## HousePrices: Final Model Summary

- Final predictors kept:

`r house.predictors`

- Removed from full model:

`r house.removed.text`

- The final model keeps the variables that most directly describe usable
  house characteristics and preferred location.
- Because no term drops from the full model, the main housing features
  appear to add complementary signal rather than redundant overlap.

```{r}
kable(house.summary, row.names = FALSE)
```

---

class: tiny-table compact small-title

## HousePrices: Final Model Terms

```{r}
kable(house.key, row.names = FALSE)
```

- Larger lots, more stories and bathrooms, air conditioning, a garage,
  and preferred location all move fitted price upward when the other
  features are held fixed.
- The coefficient table shows that the strongest evidence sits in
  structural and amenity terms rather than in one single dominant
  feature.

---

class: plot-slide small-table

## HousePrices: Diagnostics

```{r fig.width=9, fig.height=4.2}
par(mfrow = c(1, 2))
qqnorm(house.model.final$residuals, main = "QQ Plot")
qqline(house.model.final$residuals)
plot(
  house.model.final$fitted.values,
  house.model.final$residuals,
  pch = 16,
  xlab = "Fitted",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)
abline(h = 0, lty = "dashed")
par(mfrow = c(1, 1))
```

```{r}
kable(house.diag, row.names = FALSE)
```

- The residual checks do not show a single breakdown large enough to
  undermine the model for the main inference and prediction goals here.

---

class: plot-slide

## HousePrices: Bootstrap Coefficient Intervals

```{r fig.width=9.2, fig.height=5.2}
house.boot.plot <- subset(house.boot, term != "(Intercept)")
house.boot.plot <- house.boot.plot[nrow(house.boot.plot):1, ]

plot(
  house.boot.plot$estimate,
  seq_len(nrow(house.boot.plot)),
  xlim = range(c(house.boot.plot$lower, house.boot.plot$upper)),
  pch = 16,
  yaxt = "n",
  ylab = "",
  xlab = "Coefficient estimate",
  main = "Bootstrap 95% intervals for final-model coefficients"
)

segments(
  house.boot.plot$lower, seq_len(nrow(house.boot.plot)),
  house.boot.plot$upper, seq_len(nrow(house.boot.plot)),
  lwd = 2,
  col = "#0b6e4f"
)

abline(v = 0, lty = "dashed", col = "gray50")
axis(2, at = seq_len(nrow(house.boot.plot)), labels = house.boot.plot$term, las = 2)
```

- These are empirical 95% intervals from `200` bootstrap resamples of
  the final training set.
- Intervals far from `0` support stable coefficient direction.
- The intercept is omitted because its much larger scale would compress
  the predictor effects.
- The most practically important retained terms stay on one side of `0`,
  which supports stable sign interpretation.

---

class: small-table

## HousePrices: Prediction Examples

```{r}
kable(house.pred.table, row.names = FALSE)
```

- The confidence interval estimates the mean price.
- The prediction interval is wider because it predicts an individual
  home sale.
- The higher-amenity scenario also gets the higher fitted price, which
  matches the coefficient signs in the final model.

---

class: plot-slide

## HousePrices: CI and PI Overlay

```{r fig.width=9.5, fig.height=4.7}
plot(
  house.fit.x,
  house.train.new$price[house.order],
  pch = 16,
  col = "gray40",
  xlab = "Fitted mean response (ordered low to high)",
  ylab = "Price",
  main = "Actual data with fitted CI/PI and new prediction intervals"
)

lines(house.fit.x, house.fit.x, col = "#0b6e4f", lwd = 2)
lines(house.fit.x, house.actual.conf[house.order, "lwr"], col = "#0b6e4f", lty = 2)
lines(house.fit.x, house.actual.conf[house.order, "upr"], col = "#0b6e4f", lty = 2)
lines(house.fit.x, house.actual.pred[house.order, "lwr"], col = "#c97c10", lty = 3)
lines(house.fit.x, house.actual.pred[house.order, "upr"], col = "#c97c10", lty = 3)

points(house.new.fit, house.new.fit, pch = 17, col = "#b83227", cex = 1.3)
segments(house.new.fit, house.pred[, "lwr"], house.new.fit, house.pred[, "upr"],
         col = "#f0b44c", lwd = 2)
segments(house.new.fit, house.conf[, "lwr"], house.new.fit, house.conf[, "upr"],
         col = "#b83227", lwd = 3)
```

- Gray points are actual observed prices.
- The x-axis is fitted mean response, sorted from low to high.
- The green line is the fitted mean.
- Orange bands show PI for observed-data rows.
- Red triangles and bars now sit at their fitted locations on the green
  line.
- The wider orange bands show that individual-home prediction is much
  harder than estimating the mean response.

---

class: plot-slide

## HousePrices: Predicted vs Actual (Test Set)

```{r fig.width=9, fig.height=5.0}
plot(
  house.prediction,
  house.test$price,
  pch = 16,
  col = adjustcolor("#2a6f97", alpha.f = 0.6),
  xlab = "Predicted Price",
  ylab = "Actual Price",
  main = "Predicted vs Actual — held-out test set"
)
abline(0, 1, col = "#b83227", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = "Perfect prediction line (slope = 1)",
  col = "#b83227",
  lty = 2,
  lwd = 2,
  bty = "n"
)
```

- Points close to the dashed line indicate accurate test-set predictions.
- Deviations are largest at high price values, showing more uncertainty for expensive homes.
- The overall scatter is consistent with the test RMSE reported in the final model summary.

---

## CASchools: Dataset Snapshot

.pull-left[
- `420` observations and `9` analysis variables
- Response: average test score `score`
- `str`: students per teacher
- `calworks`: percent of students on CalWORKs
- `lunch`: percent eligible for subsidized lunch
- `income`: district income level
- `english`: percent of English learners
- `grades`: district grade-span category; one fitted indicator is
  `gradesKK-08` for K-8 districts
- Missing values: `0`
- The covariates mix school resources with student-composition measures,
  so the model is mainly explanatory rather than causal.
]

.pull-right[
```{r}
kable(
  data.frame(
    Item = c("Rows", "Columns", "Missing"),
    Value = c(nrow(schools.data), ncol(schools.data), sum(is.na(schools.data)))
  ),
  row.names = FALSE
)
```
]

---

class: plot-slide

## CASchools: Exploratory Overview

```{r fig.width=9.5, fig.height=4.8}
par(mfrow = c(1, 3), mar = c(4.5, 4, 3, 1))

hist(
  schools.data$score,
  breaks = 25,
  col = "#2a6f97",
  border = "white",
  xlab = "Score",
  main = "Distribution of Score"
)

plot(
  schools.data$lunch,
  schools.data$score,
  pch = 16,
  col = adjustcolor("#0b6e4f", alpha.f = 0.5),
  xlab = "% Subsidized Lunch",
  ylab = "Score",
  main = "Score vs Lunch"
)

plot(
  schools.data$income,
  schools.data$score,
  pch = 16,
  col = adjustcolor("#b83227", alpha.f = 0.5),
  xlab = "District Income",
  ylab = "Score",
  main = "Score vs Income"
)

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- Score is roughly symmetric and bell-shaped across districts.
- The negative relationship with lunch rate is the strongest raw signal in the data.
- Income shows a clear positive association, confirming that socioeconomic factors dominate the score pattern.

---

class: small-table

## CASchools: Interim Process

```{r}
kable(schools.process, row.names = FALSE)
```

- The main question is whether school-score variation is driven more by
  student composition than by resource variables.

---

class: plot-slide

## CASchools: CV Across the Selection Ladder

```{r fig.width=9.2, fig.height=4.7}
par(mar = c(7, 4.5, 4, 1))

boxplot(
  RMSE ~ factor(step_num, labels = schools.cv.labels),
  data = schools.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  main = "Repeated 5-fold CV on the training-set model ladder"
)

lines(seq_along(schools.cv.labels), schools.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_along(schools.cv.labels), schools.cv.summary$RMSE, pch = 16, col = "#123c55")
points(schools.cv.best.idx, schools.cv.summary$RMSE[schools.cv.best.idx], pch = 19, col = "#0b6e4f", cex = 1.2)

abline(h = schools.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "CV-best step", "AIC-selected step mean CV RMSE"),
  col = c("#123c55", "#0b6e4f", "#b83227"),
  pch = c(16, 19, NA),
  lty = c(1, NA, 2),
  lwd = c(2, NA, 2),
  bty = "n"
)

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

- The ladder is built from the training-set full model.
- Each step again summarizes `30` validation RMSE values from `5` folds
  repeated `6` times.
- This checks whether a simpler school-score model is competitive before
  touching the test set.
- Here the AIC step and CV step agree, so the red line lands at the best
  step.
- Agreement between AIC and CV makes the reduced school model much more
  defensible.

---

class: small-table

## CASchools: AIC Step vs CV Step

```{r}
kable(schools.compare.display, row.names = FALSE)
```

- The agreement here says the AIC-selected reduction is not just an
  artifact of one criterion.

---

class: small-table

## CASchools: Sensitivity to Cook Filtering

```{r}
kable(schools.sensitivity.display, row.names = FALSE)
```

- The main question is whether influential districts materially change
  the selected CV step or the full validation curve.

```{r}
kable(schools.test.display, row.names = FALSE)
```

- In the upper table, both RMSE columns are repeated-CV validation RMSE.
- In the lower table, the RMSE is computed on the held-out 20% test set
  by comparing predictions to the actual observed response values.
- When AIC and CV pick the same step, that test RMSE is shown once.
- The selected step does not move after filtering, so the main school
  result is not being driven by a few extreme districts.

---

class: small-table plot-slide

## CASchools: CV Ladder Before and After Cook Filtering

```{r fig.width=9.2, fig.height=4.6}
par(mfrow = c(1, 3), mar = c(7, 4.2, 3.5, 1))

boxplot(
  RMSE ~ factor(step_num, labels = schools.cv.labels),
  data = schools.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  ylim = schools.cv.ylim,
  main = "Before Cook filtering"
)
lines(seq_len(nrow(schools.cv.summary)), schools.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_len(nrow(schools.cv.summary)), schools.cv.summary$RMSE, pch = 16, col = "#123c55")
abline(h = schools.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "AIC-selected step"),
  col = c("#123c55", "#b83227"),
  pch = c(16, NA),
  lty = c(1, 2),
  lwd = c(2, 2),
  bty = "n",
  cex = 0.8
)

plot(
  schools.cv.summary$step_num,
  schools.cv.summary$RMSE,
  type = "b",
  pch = 16,
  lwd = 2,
  col = "gray40",
  xaxt = "n",
  xlab = "",
  ylab = "Mean validation RMSE",
  ylim = schools.cv.ylim,
  main = "Mean CV RMSE overlay"
)
lines(
  schools.clean.cv.summary$step_num,
  schools.clean.cv.summary$RMSE,
  type = "b",
  pch = 17,
  lwd = 2,
  col = "#0b6e4f"
)
abline(h = schools.aic.cv.rmse, col = "gray40", lty = 2, lwd = 2)
abline(h = schools.clean.aic.cv.rmse, col = "#0b6e4f", lty = 2, lwd = 2)
axis(1, at = schools.cv.summary$step_num, labels = schools.cv.labels, las = 2, cex.axis = 0.8)
legend(
  "topleft",
  legend = c("Before Cook", "After Cook", "Before AIC RMSE", "After AIC RMSE"),
  col = c("gray40", "#0b6e4f", "gray40", "#0b6e4f"),
  pch = c(16, 17, NA, NA),
  lty = c(1, 1, 2, 2),
  lwd = c(2, 2, 2, 2),
  bty = "n",
  cex = 0.8
)

boxplot(
  RMSE ~ factor(step_num, labels = schools.clean.cv.labels),
  data = schools.clean.cv,
  las = 2,
  cex.axis = 0.8,
  xlab = "",
  ylab = "Validation RMSE",
  ylim = schools.cv.ylim,
  main = "After Cook filtering"
)
lines(seq_len(nrow(schools.clean.cv.summary)), schools.clean.cv.summary$RMSE, col = "#123c55", lwd = 2)
points(seq_len(nrow(schools.clean.cv.summary)), schools.clean.cv.summary$RMSE, pch = 16, col = "#123c55")
abline(h = schools.clean.aic.cv.rmse, col = "#b83227", lty = 2, lwd = 2)
legend(
  "topleft",
  legend = c("Mean CV RMSE", "AIC-selected step"),
  col = c("#123c55", "#b83227"),
  pch = c(16, NA),
  lty = c(1, 2),
  lwd = c(2, 2),
  bty = "n",
  cex = 0.8
)

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- The best step stays the same after Cook filtering, but the full curve
  shifts down, which makes the selected model look more stable.
- That means filtering helps fit quality, but it does not change which
  predictors matter.

---

class: tiny-table compact

## CASchools: Final Model

- Final predictors kept:

`r schools.predictors`

- Removed from full model:

`r schools.removed.text`

.pull-left[
```{r}
kable(schools.summary, row.names = FALSE)
```
]

.pull-right[
```{r}
kable(schools.key, row.names = FALSE)
```
]

- `gradesKK-08` is kept in the pooled model because it captures a real
  district-type difference; dropping it would likely push that effect
  into the other coefficients.
- The district-type extensions test that choice directly and find little
  evidence that district type changes the slopes of the shared
  predictors.
- Once poverty proxies and district type are in the model, extra
  resource variables add much less explanatory power.

---

class: plot-slide small-table

## CASchools: Diagnostics

```{r fig.width=9, fig.height=4.2}
par(mfrow = c(1, 2))
qqnorm(schools.model.final$residuals, main = "QQ Plot")
qqline(schools.model.final$residuals)
plot(
  schools.model.final$fitted.values,
  schools.model.final$residuals,
  pch = 16,
  xlab = "Fitted",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)
abline(h = 0, lty = "dashed")
par(mfrow = c(1, 1))
```

```{r}
kable(schools.diag, row.names = FALSE)
```

- The residual diagnostics are acceptable for this project; the larger
  interpretive question is coefficient meaning, not obvious model
  failure.

---

class: plot-slide

## CASchools: Bootstrap Coefficient Intervals

```{r fig.width=9.2, fig.height=5.2}
schools.boot.plot <- subset(schools.boot, term != "(Intercept)")
schools.boot.plot <- schools.boot.plot[nrow(schools.boot.plot):1, ]

plot(
  schools.boot.plot$estimate,
  seq_len(nrow(schools.boot.plot)),
  xlim = range(c(schools.boot.plot$lower, schools.boot.plot$upper)),
  pch = 16,
  yaxt = "n",
  ylab = "",
  xlab = "Coefficient estimate",
  main = "Bootstrap 95% intervals for final-model coefficients"
)

segments(
  schools.boot.plot$lower, seq_len(nrow(schools.boot.plot)),
  schools.boot.plot$upper, seq_len(nrow(schools.boot.plot)),
  lwd = 2,
  col = "#0b6e4f"
)

abline(v = 0, lty = "dashed", col = "gray50")
axis(2, at = seq_len(nrow(schools.boot.plot)), labels = schools.boot.plot$term, las = 2)
```

- These are empirical 95% intervals from `200` bootstrap resamples.
- A log scale would not work here because several coefficients are
  negative.
- The intercept is omitted so the predictor intervals do not collapse
  visually toward `0`.
- `lunch` and `income` remain the most stable effects, while weaker
  terms sit closer to `0` and deserve more caution.

---

class: small-table

## CASchools: Prediction Examples

```{r}
kable(schools.pred.table, row.names = FALSE)
```

- Lower lunch rates and higher district income lead to higher predicted
  scores in the final model.
- The wider prediction intervals again show that district-level mean
  structure is estimated more tightly than individual-district outcome.

---

class: plot-slide

## CASchools: CI and PI Overlay

```{r fig.width=9.5, fig.height=4.7}
plot(
  schools.fit.x,
  schools.train.new$score[schools.order],
  pch = 16,
  col = "gray40",
  xlab = "Fitted mean response (ordered low to high)",
  ylab = "Score",
  main = "Actual data with fitted CI/PI and new prediction intervals"
)

lines(schools.fit.x, schools.fit.x, col = "#0b6e4f", lwd = 2)
lines(schools.fit.x, schools.actual.conf[schools.order, "lwr"], col = "#0b6e4f", lty = 2)
lines(schools.fit.x, schools.actual.conf[schools.order, "upr"], col = "#0b6e4f", lty = 2)
lines(schools.fit.x, schools.actual.pred[schools.order, "lwr"], col = "#c97c10", lty = 3)
lines(schools.fit.x, schools.actual.pred[schools.order, "upr"], col = "#c97c10", lty = 3)

points(schools.new.fit, schools.new.fit, pch = 17, col = "#b83227", cex = 1.3)
segments(schools.new.fit, schools.pred[, "lwr"], schools.new.fit, schools.pred[, "upr"],
         col = "#f0b44c", lwd = 2)
segments(schools.new.fit, schools.conf[, "lwr"], schools.new.fit, schools.conf[, "upr"],
         col = "#b83227", lwd = 3)
```

- The x-axis is fitted mean response, sorted from low to high.
- New prediction cases now sit at their own fitted locations instead of
  being appended at the end.
- The spread around the fitted line is consistent with the wider orange
  prediction bands for new districts.

---

class: plot-slide

## CASchools: Predicted vs Actual (Test Set)

```{r fig.width=9, fig.height=5.0}
plot(
  schools.prediction,
  schools.test$score,
  pch = 16,
  col = adjustcolor("#0b6e4f", alpha.f = 0.6),
  xlab = "Predicted Score",
  ylab = "Actual Score",
  main = "Predicted vs Actual — held-out test set"
)
abline(0, 1, col = "#b83227", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = "Perfect prediction line (slope = 1)",
  col = "#b83227",
  lty = 2,
  lwd = 2,
  bty = "n"
)
```

- The tight scatter around the diagonal shows the model generalizes well to unseen districts.
- Residual spread is roughly even across score levels, consistent with the Breusch-Pagan result.
- Any remaining deviations reflect unexplained district-level variation, not a systematic model failure.

---

class: center, middle, inverse

# Part II

## Finance Application

<span class="accent">6 NYSE stocks, 3 years of close prices, and S&P 500 benchmarking</span>

---

class: small-table compact

## Finance: Method Steps

```{r}
kable(finance.process, row.names = FALSE)
```

- Returns, not raw prices, are the right object for comparison because
  the stocks trade at very different price levels.
- The finance results answer different questions: average performance,
  market linkage, regression exposure, and time variation.

---

class: small-table

## Finance: Stock Selection

```{r}
kable(finance.stocks, row.names = FALSE)
```

- All six stocks are NYSE-listed.
- Industries are intentionally different to avoid choosing five similar
  firms.
- Benchmark: `^GSPC` for the S&P 500.
- The cross-sector mix gives the portfolio visibly different beta and
  correlation behavior instead of five nearly identical names.

---

class: plot-slide

## Finance: Daily Return Distributions

```{r fig.width=9.5, fig.height=5.0}
par(mfrow = c(2, 3), mar = c(4, 3.5, 3, 1))

for (ticker in finance.stocks$Ticker) {
  hist(
    finance.ret.mat[, ticker],
    breaks = 40,
    col = "#2a6f97",
    border = "white",
    xlab = "Daily Return",
    main = ticker,
    cex.main = 1
  )
}

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- All six stocks show fat-tailed return distributions — extreme days occur more often than a normal model predicts.
- The spread of each histogram confirms the annualized risk ordering shown in the return-risk plot.
- Comparing widths visually separates the more volatile names from the steadier defensive stocks.

---

class: plot-slide

## Finance: Close Prices

```{r fig.width=9.5, fig.height=5.2}
par(mfrow = c(3, 2))

plot(index(finance.close), as.numeric(finance.close[, "JPM"]),
     type = "l", xlab = "Date", ylab = "Close", main = "JPM")

plot(index(finance.close), as.numeric(finance.close[, "JNJ"]),
     type = "l", xlab = "Date", ylab = "Close", main = "JNJ")

plot(index(finance.close), as.numeric(finance.close[, "XOM"]),
     type = "l", xlab = "Date", ylab = "Close", main = "XOM")

plot(index(finance.close), as.numeric(finance.close[, "WMT"]),
     type = "l", xlab = "Date", ylab = "Close", main = "WMT")

plot(index(finance.close), as.numeric(finance.close[, "CAT"]),
     type = "l", xlab = "Date", ylab = "Close", main = "CAT")

plot(index(finance.close), as.numeric(finance.close[, "BA"]),
     type = "l", xlab = "Date", ylab = "Close", main = "BA")

plot(index(finance.close), as.numeric(finance.close[, "SP500"]),
     type = "l", xlab = "Date", ylab = "Close", main = "SP500")

par(mfrow = c(1, 1))
```

- Raw close prices mainly show different scales and market regimes.
- The next comparisons switch to returns because raw dollar levels are
  not directly comparable across stocks.

---

class: plot-slide

## Finance: Annualized Return and Risk

```{r fig.width=8.8, fig.height=5.0}
plot(
  finance.annual$Annual_Risk,
  finance.annual$Annual_Return,
  xlim = range(finance.annual$Annual_Risk) + c(-0.01, 0.02),
  ylim = range(finance.annual$Annual_Return) + c(-0.02, 0.03),
  pch = 21,
  bg = "#2a6f97",
  col = "#16384f",
  cex = 1.4,
  xlab = "Annualized Risk",
  ylab = "Annual Return",
  main = "Return-Risk"
)
text(
  finance.annual$Annual_Risk,
  finance.annual$Annual_Return,
  labels = finance.annual$Ticker,
  pos = finance.annual.label.pos,
  offset = 0.55,
  cex = 0.9,
  xpd = NA
)
```

- `r finance.top.return` has the strongest annualized return, while
  `r finance.high.risk` carries the most risk.
- `r finance.low.risk` is the defensive end of the group, so the chart
  separates aggressive from steadier names.

---

class: small-table plot-slide compact small-title

## Finance: Correlation with S&P 500 vs Risk and Return

.pull-left[
```{r}
kable(finance.compare.display, row.names = FALSE)
```
]

.pull-right[
```{r fig.width=4.8, fig.height=4.4}
plot(
  finance.compare$Corr_SP500,
  finance.compare$Annual_Return,
  type = "n",
  xlim = range(finance.compare$Corr_SP500) + c(-0.08, 0.08),
  ylim = range(finance.compare$Annual_Return) + c(-0.03, 0.05),
  xlab = "Corr with SP500",
  ylab = "Annual Return",
  main = "Market correlation vs return"
)

symbols(
  finance.compare$Corr_SP500,
  finance.compare$Annual_Return,
  circles = finance.compare$Annual_Risk,
  inches = 0.2,
  add = TRUE,
  bg = adjustcolor("#2a6f97", alpha.f = 0.25),
  fg = "#2a6f97"
)

text(
  finance.compare$Corr_SP500,
  finance.compare$Annual_Return,
  labels = finance.compare$Ticker,
  pos = finance.label.pos,
  offset = 0.8,
  cex = 0.8,
  xpd = NA
)
```
]

- Horizontal position is correlation with the S&P 500, vertical position
  is annualized return, and bubble size reflects annualized risk.
- `r finance.high.corr` is the most market-linked stock here, but that
  does not automatically make it the highest-return name.
- `r finance.low.corr` helps show that lower market linkage can pair
  with a very different return-risk profile.

---

class: plot-slide

## Finance: Linking Risk-Return and Market Correlation

```{r fig.width=9.0, fig.height=5.0}
risk.x <- 0.8 + (finance.compare$Annual_Risk - min(finance.compare$Annual_Risk)) /
  diff(range(finance.compare$Annual_Risk)) * 1.0
corr.x <- 3.2 + (finance.compare$Corr_SP500 - min(finance.compare$Corr_SP500)) /
  diff(range(finance.compare$Corr_SP500)) * 1.0
ylims <- range(finance.compare$Annual_Return) + c(-0.03, 0.05)

plot(
  NA,
  xlim = c(0.4, 4.6),
  ylim = ylims,
  xaxt = "n",
  xlab = "",
  ylab = "Annual Return",
  main = "Risk-return points linked to market-correlation bubbles"
)

rect(0.55, ylims[1], 2.0, ylims[2], border = "#d9cdb7", col = adjustcolor("#f7efe1", alpha.f = 0.55))
rect(3.0, ylims[1], 4.45, ylims[2], border = "#d9cdb7", col = adjustcolor("#eef6f4", alpha.f = 0.65))
axis(1, at = c(1.275, 3.6), labels = c("Annualized Risk", "Corr with SP500"))

for (i in 1:nrow(finance.compare)) {
  segments(risk.x[i], finance.compare$Annual_Return[i], corr.x[i], finance.compare$Annual_Return[i],
           lty = 3, lwd = 1.5, col = adjustcolor("#7a8793", alpha.f = 0.8))
}

points(risk.x, finance.compare$Annual_Return, pch = 16, col = "#16384f", cex = 1.2)
symbols(
  corr.x,
  finance.compare$Annual_Return,
  circles = finance.compare$Annual_Risk,
  inches = 0.12,
  add = TRUE,
  bg = adjustcolor("#2a6f97", alpha.f = 0.25),
  fg = "#2a6f97"
)

text(risk.x, finance.compare$Annual_Return, labels = finance.compare$Ticker, pos = 2, offset = 0.6, cex = 0.8)
text(corr.x, finance.compare$Annual_Return, labels = finance.compare$Ticker, pos = 4, offset = 0.6, cex = 0.8)
```

- Dotted links connect each stock's annual return across the risk view
  and the market-correlation view.
- This makes it easier to compare whether a stock's return profile is
  lining up more with risk or with broad-market linkage.
- A stock can be very market-linked without having the most extreme risk
  position, so correlation and volatility are related but not the same.

---

class: plot-slide

## Finance: Correlation Heatmap

```{r fig.width=8.8, fig.height=5.0}
par(mar = c(6, 6, 4, 2))

image(
  1:ncol(finance.corr),
  1:nrow(finance.corr),
  t(finance.corr[nrow(finance.corr):1, ]),
  col = colorRampPalette(c("#16384f", "#fbf7ef", "#b84d32"))(100),
  zlim = c(-1, 1),
  axes = FALSE,
  xlab = "",
  ylab = "",
  main = "Return correlation heatmap"
)

axis(1, at = 1:ncol(finance.corr), labels = colnames(finance.corr), las = 2)
axis(2, at = 1:nrow(finance.corr), labels = rev(rownames(finance.corr)), las = 2)

for (i in 1:nrow(finance.corr)) {
  for (j in 1:ncol(finance.corr)) {
    text(j, nrow(finance.corr) - i + 1, labels = sprintf("%.2f", finance.corr[i, j]), cex = 0.8)
  }
}

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

- Warmer colors mean stronger positive correlation.
- Cooler colors mean weaker market co-movement.
- Correlation matters because diversification is weaker when all assets
  move together.
- The uneven color pattern shows that diversification benefit depends on
  which names are combined, not just how many names are held.

---

class: plot-slide

## Finance: Cumulative Returns

```{r fig.width=9, fig.height=4.8}
matplot(
  index(finance.cum),
  coredata(finance.cum),
  type = "l",
  lty = 1,
  lwd = 2,
  col = 1:7,
  xlab = "Date",
  ylab = "Growth of $1",
  main = "Cumulative Returns"
)

legend(
  "topleft",
  legend = colnames(finance.cum),
  col = 1:7,
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

- `r finance.best.cum` finishes with the strongest growth path, while
  `r finance.worst.cum` trails the group.
- The path differences show why average return alone can hide very
  different investment experiences.

---

class: small-table

## Finance: Alpha, Beta, and R-Squared

```{r}
kable(finance.capm.display, row.names = FALSE)
```

- `beta` measures sensitivity to market return.
- `alpha` measures abnormal return relative to the S&P 500.
- `R^2` measures how much of stock-return variation is explained by the
  market.
- `r finance.high.beta` is the most market-sensitive name here, and
  `r finance.high.rsq` is the stock whose returns are most tightly tied
  to broad market movement.

---

class: plot-slide

## Finance: CAPM Fitted vs Actual Returns

```{r fig.width=9.5, fig.height=5.0}
par(mfrow = c(2, 3), mar = c(4, 3.5, 3, 1))

for (ticker in finance.stocks$Ticker) {
  model <- lm(finance.ret.mat[, ticker] ~ finance.ret.mat[, "SP500"])
  plot(
    fitted(model),
    finance.ret.mat[, ticker],
    pch = 16,
    col = adjustcolor("#2a6f97", alpha.f = 0.35),
    cex = 0.6,
    xlab = "CAPM Fitted",
    ylab = "Actual Return",
    main = ticker,
    cex.main = 1
  )
  abline(0, 1, col = "#b83227", lwd = 1.5, lty = 2)
}

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

- Stocks with higher R-squared cluster more tightly around the diagonal; lower R-squared stocks show wider scatter.
- The spread in each panel is the stock-specific variation the CAPM market factor cannot explain.
- This complements the R-squared column in the CAPM table by making the unexplained residual variation visible.

---

class: plot-slide

## Finance: Rolling 60-Day Beta

```{r fig.width=9, fig.height=4.8}
par(mar = c(7, 4.1, 4.1, 2.1))

matplot(
  index(finance.roll.beta),
  coredata(finance.roll.beta),
  type = "l",
  lty = 1,
  lwd = 2,
  col = 1:6,
  xlab = "Date",
  ylab = "Rolling 60-day beta",
  main = "Rolling beta relative to the S&P 500"
)

legend(
  "bottom",
  inset = c(0, -0.28),
  legend = colnames(finance.roll.beta),
  col = 1:6,
  lty = 1,
  lwd = 2,
  horiz = TRUE,
  xpd = NA,
  bty = "n"
)

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

- Beta above `1` means the stock moved more aggressively than the market
  in that 60-day window.
- Beta blends co-movement with relative volatility, so a rising line can
  mean tighter market linkage, larger stock swings, or both.
- A single full-sample beta misses these shifts, so rolling beta is more
  informative about changing exposure.

---

class: plot-slide

## Finance: Rolling 60-Day Correlation

```{r fig.width=9, fig.height=4.8}
par(mar = c(7, 4.1, 4.1, 2.1))

matplot(
  index(finance.roll.corr),
  coredata(finance.roll.corr),
  type = "l",
  lty = 1,
  lwd = 2,
  col = 1:6,
  xlab = "Date",
  ylab = "Rolling 60-day correlation",
  main = "Rolling correlation with the S&P 500"
)

legend(
  "bottom",
  inset = c(0, -0.28),
  legend = colnames(finance.roll.corr),
  col = 1:6,
  lty = 1,
  lwd = 2,
  horiz = TRUE,
  xpd = NA,
  bty = "n"
)

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

- Correlation isolates co-movement strength only, without the
  relative-volatility piece.
- Together, the rolling beta and rolling correlation plots show whether
  a stock became more market-linked because it moved more in sync with
  the S&P 500, because it got more volatile, or both.
- Comparing the two rolling plots separates “moves with the market” from
  “moves more violently than the market.”

---

class: center, middle, inverse

# Final Takeaways

- `HousePrices`: the best CV step is `S1`, and Cook filtering improves
  the whole validation ladder while keeping house-price prediction
  strong.
- `CASchools`: AIC and CV agree on the same reduced step, and the final
  model links stronger scores with lower poverty measures and higher
  district income.
- Finance: the six stocks occupy different risk-return positions, show
  different market linkage to the S&P 500, and that linkage changes over
  time in the rolling beta and rolling correlation plots.

---

class: center, middle, inverse

# Appendix

CASchools district-type extensions

---

class: small-table small-title

## Appendix: CASchools Within-Group Fit Check

```{r}
kable(schools.within_group.compare.display, row.names = FALSE)
```

- These are the right `R^2` comparisons to make: pooled-model fit inside
  each district type versus a separate model fit inside that same
  district type.
- Splitting helps `KK-06` by about `r sprintf('%.4f', kk06.gain)` and
  changes `KK-08` by only `r sprintf('%.4f', kk08.gain)`.
- So the pooled model was already fitting both district types fairly
  well; stratifying is not a dramatic upgrade.

---

class: tiny-table compact small-title

## Appendix: CASchools Stratified Coefficients and P Values

```{r}
kable(schools.stratified.term.compact, row.names = FALSE)
```

- `lunch` and `income` are stable in both district types.
- `english` is clearly significant in `KK-08`, but only marginal in the
  smaller `KK-06` sample.
- `str` stays negative in both groups, but its significance is much less
  stable than the socioeconomic measures.

---

class: tiny-table compact small-title

## Appendix: CASchools Interaction Slopes by District Type

```{r}
kable(schools.interaction.anova.display, row.names = FALSE)
```

```{r}
kable(schools.interaction.slope.compact, row.names = FALSE)
```

- In R's treatment coding, `KK-06` is the baseline group, so the raw
  interaction coefficients are slope differences rather than full
  `KK-08` slopes.
- Reconstructing both group slopes shows the same main pattern in each
  district type: `lunch` stays negative and `income` stays positive.
- The joint F-test and the large delta p-values both say the slope
  differences are weak.

---

class: plot-slide compact small-title

## Appendix: CASchools Interaction Plot Explained

```{r fig.width=8.8, fig.height=4.4}
plot(
  range(schools.interaction.grid$income),
  range(schools.interaction.grid$fit),
  type = "n",
  xlab = "Income",
  ylab = "Predicted score",
  main = "Predicted score vs income, holding other predictors at medians"
)

for (g in schools.strata) {
  sub <- schools.interaction.grid[schools.interaction.grid$grades == g, ]
  lines(
    sub$income,
    sub$fit,
    lwd = 2,
    col = ifelse(g == "KK-06", "#b83227", "#123c55")
  )
}

legend(
  "topleft",
  legend = schools.strata,
  col = c("#b83227", "#123c55"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

- The joint F-test gives a p-value of `r schools.interaction.p.value`,
  so adding all five interactions does not materially improve fit.
- Only `income` varies here; the other predictors are fixed at their
  sample medians, and the near-parallel lines show district type mainly
  shifts the fitted level rather than the slope pattern.
